---
title: "Documenting Software Used"
description: "Automate documenting the software used to generate an output file."
date: "2026-09-30"
date-modified: "2026-09-30"
draft: false
categories:
  - Quarto Tips
  - Reproducible Research
  - Automation Principle
  - Preservation Principle
execute:
  echo: true
---

This post is about how I document the software environment used to generate 
output files. 

# Why Documenting the Software Used is Useful
The documentation criterion for reproducibility requires telling other people 
what software environment you used to conduct analyses. The most reliable way to 
reproduce your work is to run your code, on your data, with the exact same
software that you used. It is not enough to just say that you used R or a
specific set of R packages because both the base R system and R packages change
over time. Some package authors publish new versions quite frequently.

Changes between versions can fix (or introduce) bugs and add, change, or remove 
functions, arguments, syntax, interfaces, or dependencies on other software. So 
someone else may be unable to reproduce your work even though they are using the 
same set of packages if they have either older or newer versions than you were
using. 

Therefore, you need to record the specific version numbers of all the software 
used for a given analysis. Doing so has helped me solve issues when routine R
package updates broke scripts that previously worked reliably because I was able 
to systematically track down which pieces of my software environment had changed
then either revert to using an older version of a specific package or update my
code to work with the newer version. Naturally, this all works better if you 
[keep date-stamped production outputs](https://sjpierce.github.io/posts/output/draft_vs_production.html). 

# Automation Helps
I run package updates frequently and often use a large collection of R packages
to manage data and run analyses. Loading an R package also means other packages 
that it depends on also get loaded. Manually writing out the version numbers for
all of them would be tedious and error-prone. The solution is to automate
capturing that information by adding a few function calls at the end of scripts.

So, I often include a relatively standardized Software Information section and 
code chunks like those shown below in my scripts.

# Load R Packages 
Near the beginning of my Quarto files, I'll include a chunk to load R packages. 
The ones loaded below offer functions I use to automate documenting my software 
environment. 

```{r}
#| label: load-packages
#| message: false

library(devtools)  # for session_info()
library(piercer)   # for git_report(), which_latex().
library(quarto)    # for quarto_version()
library(rmarkdown) # for pandoc_version()
```

# Software Information
I used [R](https://www.r-project.org/) as the main computing environment and 
[Quarto](https://quarto.org/) scripts to enhance reproducibility. I used 
[RStudio](www.rstudio.org) as the editor to interface with R and Quarto. 

- PDF software chain:
  **qmd file > RStudio > Quarto > R > knitr > md file > Pandoc > tex file > TinyTeX > PDF file**.
- HTML software chain:
  **qmd file > RStudio > Quarto > R > knitr > md file > Pandoc > HTML file**.
- [Quarto `{r} quarto_version()`](https://quarto.org/) runs `*.qmd` files 
  through [R](https://www.r-project.org/) and [knitr](https://yihui.org/knitr/) 
  to produce `*.md` markdown files.
- [Pandoc `{r} rmarkdown::pandoc_version()`](https://pandoc.org) converts 
  markdown files (`*.md`) to other formats, including LaTeX (`*.tex`) and HTML 
  (`*.html`) among others.
- [TinyTeX](https://yihui.org/tinytex/) compiles LaTeX files (`*.tex`) into PDF 
  files. 

## Versions
This document was generated using the following computational environment and 
dependencies: 

``` {r}
#| label: show-version

# Quarto (used inline above in the bullet list)
quarto_version()

# Pandoc (used inline above in the bullet list) 
pandoc_version()

# TinyTex or other LaTeX software. 
which_latex()

# Get R and R package version numbers in use.
devtools::session_info()
```

## Git Details
The current Git commit details and status are:

``` {r}
#| label: git-details
#| cfsize: scriptsize

git_report()
```

This is useful because it tells us exactly which commit in the Git history 
we would need to be using to make sure we are running the exact same code. 
Sometimes another person is not using the most current code, or has changed the 
code in some way since it was last committed.  

::: {.callout-tip}
* Untracked files are files located in the repository that Git has not been told 
  to entirely ignore, but have also not been committed into the version history. 
* Unstaged changes to files indicate that some of the contents have been 
  modified since the last time the file was committed to Git. In production 
  runs, we want the Git output to not show any unstaged changes to key files!

::: 
